---
title: "Weather Analysis"
author: "Martin Connolly"
date: "June 21, 2015"
output: html_document
---

# Synopsis

This report uses [data](https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2) from the U.S. National Oceanic and Atmospheric Administration's (NOAA) storm database to answer 2 questions;

1. Across the United States, which types of events are most harmful with respect to population health?

2. Across the United States, which types of events have the greatest economic consequences?

The key findings are that more lives could be saved by addressing tornado preparedness. Possibly public awareness campaigns about steps to take in the event of a tornado warning and that more property and crop damage could be reduced by increasing flood defenses.


# Data Processing

## Loading and preprocessing the data


```{r, echo=FALSE}
# set working directory
dir_name <- "C:/Users/martin.connolly/Documents/GitHub/RepData_PeerAssignment2"
setwd(dir_name)
```

```{r, echo=TRUE, message=FALSE, warning=FALSE}
require(dplyr, quietly = TRUE, warn.conflicts = FALSE)
require(ggplot2, quietly = TRUE, warn.conflicts = FALSE)
```

```{r, echo=TRUE, cache=TRUE}

storm_data <- read.csv(bzfile("data/repdata_data_StormData.csv.bz2"),
                       head = TRUE,
                       stringsAsFactors = TRUE)
```

We need to take into account the values in the CROPDMGEXP and PROPDMGEXP, which determine if the CROPDMG and PROPDMG variables are in thousands, millions or billions.


```{r, echo=TRUE, cache=TRUE}
subset_storm_data <- select(storm_data, c(EVTYPE, FATALITIES, INJURIES, PROPDMG, PROPDMGEXP, CROPDMG, CROPDMGEXP))

subset_storm_data$CROP_MULT[subset_storm_data$CROPDMGEXP %in% c("k", "K")] <- 1000
subset_storm_data$CROP_MULT[subset_storm_data$CROPDMGEXP %in% c("m", "M")] <- 1000000
subset_storm_data$CROP_MULT[subset_storm_data$CROPDMGEXP %in% c("b", "B")] <- 1000000000
subset_storm_data$CROP_MULT[subset_storm_data$CROPDMGEXP == "2"] <- 100
subset_storm_data$CROP_MULT[is.na(subset_storm_data$CROP_MULT)] <- 1

subset_storm_data$PROP_MULT[subset_storm_data$PROPDMGEXP %in% c("k", "K", "3")] <- 1000
subset_storm_data$PROP_MULT[subset_storm_data$PROPDMGEXP %in% c("m", "M", "6")] <- 1000000
subset_storm_data$PROP_MULT[subset_storm_data$PROPDMGEXP %in% c("b", "B")] <- 1000000000
subset_storm_data$CROP_MULT[subset_storm_data$CROPDMGEXP == "1"] <- 10
subset_storm_data$CROP_MULT[subset_storm_data$CROPDMGEXP == "2"] <- 100
subset_storm_data$CROP_MULT[subset_storm_data$CROPDMGEXP == "4"] <- 10000
subset_storm_data$CROP_MULT[subset_storm_data$CROPDMGEXP == "5"] <- 100000
subset_storm_data$CROP_MULT[subset_storm_data$CROPDMGEXP == "7"] <- 10000000
subset_storm_data$CROP_MULT[subset_storm_data$CROPDMGEXP == "8"] <- 100000000
subset_storm_data$PROP_MULT[is.na(subset_storm_data$PROP_MULT)] <- 1

subset_storm_data <- mutate(subset_storm_data, CROPDMG = CROPDMG * CROP_MULT)
subset_storm_data <- mutate(subset_storm_data, PROPDMG = PROPDMG * PROP_MULT)
```

#Results

We can see below the top 5 events for injuries and fatalities and the top 5 events for damage to crops and property


```{r, echo = TRUE}
grouped_data <- subset_storm_data %>% group_by(EVTYPE)
health_effects <- arrange(summarise(grouped_data, TOTAL = sum(FATALITIES) + sum(INJURIES)),
                          desc(TOTAL))

head(health_effects, 5)

ggplot(health_effects[1:5,], aes(x = EVTYPE, y = TOTAL)) + geom_bar(stat = "identity") +
        ylab("Injuries and Fatalities") + xlab("Event Type") + ggtitle("Top 5 injuries and fatalities by event")

economic_effects <- arrange(summarise(grouped_data, TOTAL = sum(CROPDMG) + sum(PROPDMG)),
                          desc(TOTAL))

head(economic_effects, 5)

ggplot(economic_effects[1:5,], aes(x = EVTYPE, y = TOTAL)) + geom_bar(stat = "identity") +
        ylab("Crop and Property Damage") + xlab("Event Type") + ggtitle("Top 5 estimated economic effects by event")
```

